### 2020-06-04 writeup/intro.tex
intro wip

\vspace{-0.3in}
\begin{abstract}
\vspace{-0.1in}
%When does multi-task learning provide positive (or negative) information transfer, compared to single-task learning?
	Multi-task learning is a powerful approach in many applications such as image and text classification.
	Yet, there is little rigorous understanding of when multi-task learning performs well, even compared to single-task learning.
	To be precise, we say that multi-task learning provides \textit{positive transfer} if it outperforms single-task learning, or \textit{negative transfer} otherwise.
	In this work, we consider three components including \textit{task similarity}, \textit{sample size} and \textit{covariate shift} to determine positive vs. negative transfer.
	We formally quantify each component in the setting of high-dimensional linear regression,
% the performance of predicting a particular task given multiple tasks.
%a multi-task learning architecture that uses a shared layer for all tasks and a separate output layer for each task.
	and show that the bias-variance tradeoff of the multi-task estimator determines the effect of transfer..
%	We analyze the bias-variance tradeoff by developing tight concentration results using random matrix theory.
	%we show that the type of transfer is determined by the bias-variance tradeoff of multi-task learning estimators.
	%We develop a technical tool to analyze the.
%	We develop a technical tool on the inverse of the sum of two random matrices that naturally arises from the analysis.
	Then, we show that the performance of multi-task learning can be negatively affected as a result of: (a) decreased task similarity;
	(b) increased source sample size; (c) covariate shift under increased source sample size.
	We validate the above results on text classification tasks.
	Inspired by our theory, we show that single-task performance can help understand multi-task performance, and propose an incremental training scheduler for improving the efficiency of multi-task training.% for predicting a particular task.
%	We validate our theory and our proposed method .
\end{abstract}

\vspace{-0.1in}
\section{Introduction}

%Multi-task learning is an inductive learning mechanism to improve generalization performance using related task data.
%Many state-of-the-art results in computer vision and natural language processing are obtained using multi-task learning.
Multi-task learning is a powerful paradigm for improving performance on prediction tasks in computer vision \cite{chexnet17,ZSSGM18}, natural language processing \cite{GLUE,superglue} and many other areas \cite{ZY17}.
%In multi-task learning, having related task data is fundamental to its performance.
%Multi-task learning is particularly powerful when there is limited labeled data for a task to be solved, meanwhile more labeled data from different but related tasks is available.
%By combining multiple information sources, it is possible to share all the information in the same model.
In many settings, multiple data sources are available for predicting a particular target task.
%For example, many applications in , and many other areas have been achieved by learning from multiple tasks together.
The performance of multi-task learning depends on the relationship between the data sources and the target \cite{C97}.
When the data sources are relatively different from each other, negative transfer has often been observed \cite{PY09,AP16,BS17}.
While many empirical approaches have been proposed to mitigate negative transfer \cite{ZY17}, a precise understanding of when negative transfer occurs remains elusive in the literature \cite{R17}.
%This phenomenon, known as \textit{negative transfer}, is fundamental to the understanding of multi-task learning.
In this work, we provide a rigorous study to understand when positive or transfer occurs in the setting of high-dimensional linear regression.
We show that the bias-variance tradeoff of multi-task learning estimators determines the transfer effect, and we develop new concentration bounds to analyze the tradeoff.
%Inspired by the theory, we propose an incremental training schedule to improve multi-task training.
%We consider a setting where the target task has limited labeled data and show
%On the other hand, unless the structures across task data are well-understood, applying multi-task learning on several different datasets often result in suboptimal models (or negative transfer in more technical terms).

Understanding positive vs. negative transfer requires developing generalization bounds that scale tightly with properties of the data such as sample size.
This presents a technical challenge in the multi-task setting due to the presence of different task features, even for two tasks.
For Rademacher complexity or VC-based techniques, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.
Without a tight lower bound for the multi-task setting, it is unclear how to compare its performance to single-task learning.
Furthermore, developing a better understanding of how multi-task learning performs in terms of properties of the data is of practical importance, since a major application of multi-task learning is for settings where a particular task has limited amount of labeled data \cite{RH19}.
%For example, the sample sizes of all tasks are often assumed to be equal \cite{B00,LPTV09,LPVT11}.
%On the other hand, uneven sample sizes (or dominating tasks) have been empirically observed to cause negative transfer \cite{YKGLHF20}.
%The benefit of learning multi-task representations has also been studied for certain half-spaces \cite{} and sparse regression \cite{}.
%When all tasks are sufficiently similar, adding more labeled data improves the generalization performance for predicting a particular task \cite{WZR20}.

%\textbf{Setup and Main Results.}
In this work, we apply recent development in the random matrix theory literature \cite{erdos2017dynamical,isotropic,Anisotropic} to show how to determine positive vs. negative transfer in the setting of high-dimensional linear regression \cite{HMRT19,BLLT20}.
This setting allows us to measure the difference of task data by \textit{task similarity}, \textit{sample size} and \textit{covariate shift} precisely.
We define each component for two tasks and our formulation applies more generally.
We call the first task a source task and the second a target task.
\squishlist
	\item \textbf{Task similarity:} Assume that each task follows a linear model with parameters $\beta_i\in\real^p$, for $i=1, 2$.
	Hence, we measure the distance between the two tasks by $\norm{\beta_1 - \beta_2}$.
	\item \textbf{Sample size:} Let $n_1 = \rho_1 \cdot p$ and $n_2 = \rho_2 \cdot p$ denote the sample size of each task, where both $\rho_1, \rho_2 > 1$ are fixed values that do not grow with $p$.
	Hence, the source/target sample ratio is measured by $n_1 / n_2 = \rho_1 / \rho_2$.
	Importantly, $\rho_2$ can be a small constant (say $2$) to capture the need for more labeled data.
	\item \textbf{Covariate shift:} We assume that for every task, its features are random vectors with covariance matrix $\Sigma_1,\Sigma_2\in\real^{p\times p}$.
	%$x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and is a positive semidefinite matrix.
	Hence, we can measure covariate shift by $\Sigma_1^{1/2}\Sigma_2^{-1/2}$.
\squishend

@@

%Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
%The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We focus on the hard parameter sharing architecture with a linear shared layer for all tasks and a separate prediction head for each task \cite{R17,WZR20,MTDNN19}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_prelim}.
We revisit the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$.
Interestingly, we observe that the variance of the multi-task estimator is always smaller than the variance of the single-task estimator, because of an increased sample size.
The bias of the multi-task estimator results in a negative effect caused by the difference between $\beta_t$ and the rest of $\set{\beta_i}_{i=1}^{t-1}$.
Hence, the bias-variance tradeoff determines whether we observe positive or negative transfer.
For the setting of two tasks, we show how the variance of the multi-task estimator  scales with sample size and covariate shift in the following result.
%\textit{Our first contribution} is to develop a concentration bound that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$

>>>>

\vspace{-0.3in}
\begin{abstract}
\vspace{-0.05in}
%When does multi-task learning provide positive (or negative) information transfer, compared to single-task learning?
	Multi-task learning is a powerful approach in many applications such as image and text classification.
	Yet, there is little rigorous understanding of when multi-task learning performs well, even compared to single-task learning.
	To be precise, we say that multi-task learning provides \textit{positive transfer} if it outperforms single-task learning, or \textit{negative transfer} otherwise.
	In this work, we consider three components including \textit{task similarity}, \textit{sample size} and \textit{covariate shift} to determine positive vs. negative transfer.
	We formally quantify each component in the setting of high-dimensional linear regression,
% the performance of predicting a particular task given multiple tasks.
%a multi-task learning architecture that uses a shared layer for all tasks and a separate output layer for each task.
	and show that the bias-variance tradeoff of the multi-task estimator determines the effect of transfer..
%	We analyze the bias-variance tradeoff by developing tight concentration results using random matrix theory.
	%we show that the type of transfer is determined by the bias-variance tradeoff of multi-task learning estimators.
	%We develop a technical tool to analyze the.
%	We develop a technical tool on the inverse of the sum of two random matrices that naturally arises from the analysis.
	Then, we show that the performance of multi-task learning can be negatively affected as a result of: (a) decreased task similarity;
	(b) increased source sample size; (c) covariate shift under increased source sample size.
	We validate the above results on text classification tasks.
	Inspired by our theory, we show that single-task performance can help understand multi-task performance, and propose an incremental training scheduler for improving the efficiency of multi-task training.% for predicting a particular task.
%	We validate our theory and our proposed method .
\end{abstract}

\vspace{-0.15in}
\section{Introduction}

%Multi-task learning is an inductive learning mechanism to improve generalization performance using related task data.
%Many state-of-the-art results in computer vision and natural language processing are obtained using multi-task learning.
Multi-task learning is a powerful approach for improving performance on prediction tasks in computer vision \cite{chexnet17,ZSSGM18}, natural language processing \cite{GLUE,superglue} and many other areas \cite{ZY17}.
%In multi-task learning, having related task data is fundamental to its performance.
%Multi-task learning is particularly powerful when there is limited labeled data for a task to be solved, meanwhile more labeled data from different but related tasks is available.
%By combining multiple information sources, it is possible to share all the information in the same model.
In many settings, multiple data sources are available for predicting a particular target task.
%For example, many applications in , and many other areas have been achieved by learning from multiple tasks together.
The performance of multi-task learning depends on the relationship between the data sources and the target \cite{C97}.
When the data sources are relatively different from each other, negative transfer has often been observed \cite{PY09,AP16,BS17}.
While many empirical approaches have been proposed to mitigate negative transfer \cite{ZY17}, a precise understanding of when negative transfer occurs remains elusive in the literature \cite{R17}.
%This phenomenon, known as \textit{negative transfer}, is fundamental to the understanding of multi-task learning.
In this work, we provide a rigorous study to understand when positive or transfer occurs in the setting of high-dimensional linear regression.
We show that the bias-variance tradeoff of multi-task learning estimators determines the transfer effect, and we develop new concentration bounds to analyze the tradeoff.
%Inspired by the theory, we propose an incremental training schedule to improve multi-task training.
%We consider a setting where the target task has limited labeled data and show
%On the other hand, unless the structures across task data are well-understood, applying multi-task learning on several different datasets often result in suboptimal models (or negative transfer in more technical terms).

Understanding positive versus negative transfer requires developing generalization bounds that scale tightly with properties of each task data such as its sample size.
This presents a technical challenge in the multi-task setting due to the presence of different task features, even for two tasks.
For Rademacher complexity or VC-based techniques, the generalization error scales down as the sample sizes of all tasks increase, when applied to the multi-task setting \cite{B00,AZ05,M06,MPR16,WZR20}.
Without a tight lower bound for multi-task learning, comparing its performance to single-task learning results in vacuous bounds.
From a practical standpoint, developing a better understanding of multi-task learning in terms of properties of task data is important because it is often applied to settings in which a particular task has limited amount of labeled data \cite{RH19}.
%For example, the sample sizes of all tasks are often assumed to be equal \cite{B00,LPTV09,LPVT11}.
%On the other hand, uneven sample sizes (or dominating tasks) have been empirically observed to cause negative transfer \cite{YKGLHF20}.
%The benefit of learning multi-task representations has also been studied for certain half-spaces \cite{} and sparse regression \cite{}.
%When all tasks are sufficiently similar, adding more labeled data improves the generalization performance for predicting a particular task \cite{WZR20}.

%\textbf{Setup and Main Results.}
In this work, we apply recent development from the random matrix theory literature \cite{erdos2017dynamical,isotropic,Anisotropic} to determine transfer effect in the high-dimensional linear regression setting \cite{HMRT19,BLLT20}.
This setting allows us to measure the difference of task data by three components, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
To be precise, we define each component for two tasks below and our definition applies more generally.
We refer to the first task as the source and the second as the target.
\squishlist
	\item \textbf{Task similarity:} Assume that both tasks follow a linear model with parameters $\beta_1, \beta_2\in\real^p$, respectively.
	We measure the distance between them by $\norm{\beta_1 - \beta_2}$.
	\item \textbf{Sample size:} Let $n_1 = \rho_1 \cdot p, n_2 = \rho_2 \cdot p$ be the sample size of each task, where $\rho_1, \rho_2$ are both fixed values that do not grow with $p$.
	We measure the source/target sample ratio by $\rho_1 / \rho_2$.
%	Importantly, $\rho_2$ can be a small constant (say $2$) to capture the need for more labeled data.
	\item \textbf{Covariate shift:} Assume that the task features are random vectors with positive semidefinite covariance matrix $\Sigma_1,\Sigma_2\in\real^{p\times p}$, respectively.
	%$x = \Sigma_i^{1/2}z$, where $z\in\real^p$ consists of i.i.d. entries with mean zero and unit variance, and is a positive semidefinite matrix.
	We define covariate shift as the matrix $\Sigma_1^{1/2}\Sigma_2^{-1/2}$.
\squishend

@@

%Let $n_i = \rho_i \cdot p$ denote the data size and $X_i\in\real^{n_i\times p}$ denote the features of task $i$.
%The labels of task $i$ are given by $Y_i = X_i\beta_i + \varepsilon_i$, where $\beta_i\in\real^p$ denotes task $i$'s ground truth parameters and $\varepsilon_i$ denotes i.i.d. random noise with mean zero and variance $\sigma^2$.
We consider a multi-task estimator obtained using a shared linear layer for all tasks and a separate output layer for each task \cite{WZR20}.
This two layer model is inspired by a commonly used idea of hard parameter sharing in multi-task learning \cite{R17,MTDNN19}.
%This corresponds to minimizing $ \sum_{i=1}^t \norm{X_i B W_i - Y_i}^2$.
%Let $\hat{\beta}_t^{\MTL}$ denote the optimal multi-task estimator for the target task, which is defined precisely in Section \ref{sec_prelim}.
We revisit the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$.
Interestingly, we observe that the variance of the multi-task estimator is always smaller than the variance of the single-task estimator, because of an increased sample size.
The bias of the multi-task estimator results in a negative effect caused by the difference between $\beta_t$ and the rest of $\set{\beta_i}_{i=1}^{t-1}$.
Hence, the bias-variance tradeoff determines whether we observe positive or negative transfer.
For the setting of two tasks, we show how the variance of the multi-task estimator  scales with sample size and covariate shift in the following result.
%\textit{Our first contribution} is to develop a concentration bound that arises naturally from the bias-variance tradeoff of $\hat{\beta}_t^{\MTL}$ for two tasks.
%Let $\hat{\beta}_t^{\STL}$ denote the single-task estimator.
%Without loss of generality, let the $t$-th task denote the target task.
%Importantly, the target task's data size is a fixed constant times $p$ in the high-dimensional setting.
%Hence adding more labeled data can help improve its test performance.
%$B\in\real^{p\times r}$
%$\set{W_i \in \real^{r}}_{i=1}^t$
